The Abundance of Bullet-like Groups in $\Lambda$CDM

### 2014-04-15 paper/ms.tex
submitted version (figures in pdf)
updates for the referee

\documentclass{emulateapj}
%\documentclass{aastex}
\submitted{{\it Submitted for publication in ApJL}}
\usepackage {graphicx}

\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{graphics}
\usepackage{epsfig}  
\usepackage{float}
\bibliographystyle{apj}
\def\be{\begin{equation}}
\def\ee{\end{equation}}
\def\ba{\begin{eqnarray}}
\def\ea{\end{eqnarray}}

\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}
\newcommand{\bullb}{MACS J0025.4-1222}
\newcommand{\bulla}{1E0657---56} 
\newcommand{\bullg}{SL2S J08544-0121}
\shorttitle{Bullet Groups}
\shortauthors{Fern\'andez-Trincado et al.}

@@

\title{The abundance of Bullet-Groups in $\Lambda$CDM}
\author{J. G. Fern\'andez-Trincado$^{1,2,3}$, J. E. Forero-Romero$^1$,
  G.Foex$^4$, T. Verdugo$^3$ and V. Motta$^4$} 
\affil{$^1$ Departamento de F\'{i}sica, Universidad de los Andes,
  Cra. 1 No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\ 
       $^2$ Institute Utinam, CNRS UMR6213, Universit\'e de
  Franche-Comt\'e, OSU THETA de Franche-Comt\'e-Bourgogne,
  Besan\c{c}on, France\\ 
       $^3$ Centro de Investigaciones de Astronom\'ia, AP 264,
  M\'erida 5101-A, Venezuela\\
  $^4$ Instituto de F\'isica y Astronom\'ia, Universidad de
  Valpara\'iso, Avda. Gran Breta\~na 1111, Playa Ancha, Valpara\'iso
  2360102, Chile  
}
\email{jfernandezt87@gmail.com}
\email{je.forero@uniandes.edu.co}
\begin{abstract}

@@

The Bullet Cluster (\bulla) provided a new kind of observational
evidence for the existence of dark matter
\citep{Markevitch2004,Clowe2006}. Since then it has been used to test
the Cold Dark Matter (CDM) paradigm itself by quantifying different
aspects such as the expected displacement between the dominant dark
matter and baryonic component in a $\Lambda$CDM Universe
\citep{ForeroRomero2010} or the substructure velocity required to
produce such displacement \citep{Hayashi2006, Springel2007,
  Thompson2012}. It has also been used to constrain the dark matter
particle self-interaction cross section and to explore possible
extensions to the concordance cosmological model
\citep{Farrar2007,Lee2010,Lee2012}.

Since then, other examples of Bullet-like systems have been found; 
MACS J0025.4-1222 \citep{Bradac2008}, Abell 2744 \citep{Merten2011},
DLSCL J0916.2+2951 \citep{Dawson2012}, ZwCl 1234.0+02916
\citep{Dahle2013}. Recently \citep{Gastaldello} observed a DM-gas
displacement of $87\pm 14$\hkpc and a DM-DM separation of $186\pm
30$\hkpc in \bullg, a group-like system with a total
mass $2.4\pm 0.6 \times 10^{14}\Msun$ found in the the Strong Lensing
Legacy Survey (SL2S) sample \citep{Cabanac2007,More2012}.

Groups around this mass are $\sim10$ times more abundant than clusters
in the mass range of the Bullet Cluster $>10^{15}$\hMsun. This 
should open up the possibility of finding Bullet Groups in large
numbers to test $\Lambda$CDM. However, a larger abundance of small
mass systems has to be weighted by the probability of being in a
merger and  presenting a large displacement between the DM and
gaseous components.  These two conditions (merger rates, maximum
possible displacement) are a funcion of DM halo mass in $\Lambda$CDM
cosmologies.  A detailed statistical study to estimate the DM-gas
displacements of Bullets has been performed for clusters
\citep{ForeroRomero2010} but not for lower mass systems.

@@

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250\hMpc comoving on a side. The simulation uses the ART
code \citep{Kravtsov1997} to follow the evolution of a dark matter
density field from $z=80$ to $z=0$ sampled with $2048^3$  particles. The
cosmology used  corresponds to  the spatially flat concordance model
with the following parameters:  the density parameter for matter (dark
matter and baryons) $\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the
slope of the primordial power spectrum $n=0.95$ and the amplitude of mass
density fluctuations (at redshift z$=$0) $\sigma_8=0.82$.  These
cosmological parameters are consistent with the nine-year Wilikinson
Microwave Anisotropy Probe (WMAP) results \citep{hinshaw_etal13}. A
detailed presentation of the simulation can be found in
\citet{2011ApJ...740..102K}.

@@

All the raw data used in this Letter are available through the
Multidark database \footnote{\texttt{www.multidark.org}}
\citep{Riebe2013}.  Furthermore, in order to facilitate the
reproducibility and reuse of our results we have made available all
the data and the source code available in a public
repository \footnote{\texttt{https://github.com/Fernandez-Trincado/Bullet\_Groups-2014}}.

@@

Each one of the pairs (host \& sub-halo) is considered as a
potential Bullet system and is kept for the analysis described in the
next Section  by using the sub-halos as the tracer of the sub-dominant
dark matter clump in the merging cluster, i.e. the bullets.

@@

Bullet-like configurations have two dominand dark matter structures:
the host halo and the dominant sub-halo. We describe the kinematics of
this configuration by the  position and velocity vectors of the
sub-halo in a frame of reference where the main halo is at rest; thus
$\vec{v{}}=\vec{v}_{\rm sub}-\vec{v}_{\rm host}$ and
$\vec{r{}}=\vec{r}_{\rm sub}-\vec{r}_{\rm host}$, where the subscripts $host$
and $sub$ refer to the position of the minium of potential for the
host and sub-halo in the frame of reference of the simulation,
respectively.

@@

In this Letter we work with two quantities that could be inferred
from observations of Bullet-like systems. The projected distance
between two dominant DM clumps, $d_{\rm 2D}$, and the projected distance
between the DM and the gas clumps, $d_{\rm 2D}^{\rm bar}$.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_1_V3.eps}
\end{center}
\caption{
  Integrated probability distribution for the displacement between the
  center of the host halo and its dominant sub-halo at all redshifts
  $z=0$, $0.25$, $0.5$ and $1.0$. The left panel
  shows the results in terms of the physical displacements while the
  right panel shows the displacements normalized by the virial radius
  of the host halo. The continuous (dashed) line corresponds to the halos in the
  cluster (group) sample.
  The vertical lines show the mean value and uncertainties ($186\pm
  30\hkpc$) in the  separation between the two dark matter clumps
  estimated in \citet{Gastaldello} for the \bullg. Between $40\%$
  to $60\%$ of the groups show a displacement equal or larger than
  this observational benchmark. This fraction rises to $60\%$ and
  $80\%$ in clusters.}
\label{fig:displacement}
\end{figure*}

Figure \ref{fig:displacement} presents the integrated
probability distribution for the DM-DM displacements, $d_{\rm 2D}$. The
left panel shows the displacement in physical units and the right
panel as a fraction of the virial radius of the host halo.  The panel
with the projected 2D physical displacements also shows a vertical
stripe with the estimated displacement for the Bullet-group reported
by \cite{Gastaldello}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_3_V3.eps}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$. The left
  panel corresponds to clusters and the right panel to groups. The data
  used to construct this histogram includes halos at all redshifts.}
\label{fig:geometry}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_3.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  DM-gas displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and gas. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure*}

@@

In this Letter we estimated the fraction of galaxy groups and clusters
in a $\Lambda$CDM cosmology that could present observational features
associated to a bullet-like event. This is motivated by the recent
observational results of \citep{Gastaldello} where a system (\bullg)
on the mass range $1\times 10^{14}$\hMsun\ and velocity dispersion
$650$\kms was reported to feature a displacement between its baryonic
(gas) and dark matter components.

@@

Our main result is that a fraction of $40\%$-$60\%$ of the halos
in the group sample presents DM-DM displacement equal or larger than the
observed displacement for \bullg. For halos in the cluster sample this
fraction increases to $60\%$-$80\%$. We also derived an
estimate for the DM-baryon displacement. In the group sample
$0.1\%$-$1.0\%$ of the halos show a displacement equal or larger than
the  measurements of \bullg\ by \citep{Gastaldello}; in the cluster
sample this fraction rises to $4\%$-$10\%$.

For the case of \bullg\ a fair comparison is achieved against our cluster
sample which has statistics dominated by objects of similar mass. In
a rough comparison using the observational criteria \citep{Foex2013,
  Gastaldello} we find that the relative number of bullets
(large DM-gas displacement) with respect to a general sample of
multi-modal systems (large DM-DM displacements) is consistent with
observations; both are in the range $\sim 7\%$.

Using the same criteria we find that in the simulation there there are
$\sim 6.0\times 10^{-7}$ Mpc$^{-3}$ groups similar to \bullg. This
number density is three times larger than the computed value for
clusters. This opens up a new observational possibility with surveys
such as SL2S that target a large number of groups and estimate its
multi-modal nature from lensing analysis \citep{Foex2013}. An approach
that can be further exploited with upcoming lensing surveys (e.g. with
 he {\it Euclid} satellite) and pushes for X-ray surveys with higher
sensibility: chances are larger to find bullets in systems with lower
X-ray luminosities.

>>>>

\documentclass{emulateapj}
%\documentclass{aastex}
\submitted{{\it Submitted for publication in ApJL}}
\usepackage {graphicx}
\usepackage{hyperref}
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{graphics}
\usepackage{epsfig}  
\usepackage{float}
\bibliographystyle{apj}
\def\be{\begin{equation}}
\def\ee{\end{equation}}
\def\ba{\begin{eqnarray}}
\def\ea{\end{eqnarray}}

\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}
\newcommand{\bullb}{MACS J0025.4-1222}
\newcommand{\bulla}{1E0657---56} 
\newcommand{\bullg}{SL2S J08544-0121}
\shorttitle{Bullet groups}
\shortauthors{Fern\'andez-Trincado et al.}

@@

\title{The Abundance of Bullet-like Groups in $\Lambda$CDM}
\author{J. G. Fern\'andez-Trincado$^{1,2,3}$, J. E. Forero-Romero$^1$,
  G.Foex$^4$, T. Verdugo$^3$ and V. Motta$^4$} 
\affil{$^1$ Departamento de F\'{i}sica, Universidad de los Andes,
  Cra. 1 No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\ 
       $^2$ Institute Utinam, CNRS UMR6213, Universit\'e de
  Franche-Comt\'e, OSU THETA de Franche-Comt\'e-Bourgogne,
  Besan\c{c}on, France\\ 
       $^3$ Centro de Investigaciones de Astronom\'ia, AP 264,
  M\'erida 5101-A, Venezuela\\
  $^4$ Instituto de F\'isica y Astronom\'ia, Universidad de
  Valpara\'iso, Avda. Gran Breta\~na 1111, Playa Ancha, Valpara\'iso
  2360102, Chile  
}
\email{jfernandezt87@gmail.com}
\email{je.forero@uniandes.edu.co}
\begin{abstract}

@@

The Bullet Cluster (\bulla) provided a new kind of observational
evidence for the existence of dark matter
\citep{Markevitch2004,Clowe2006}. Since then it has been used to test
the Cold Dark Matter (CDM) paradigm itself by quantifying different
aspects such as the expected displacement between the dominant dark
matter and baryonic component in a $\Lambda$CDM Universe
\citep{ForeroRomero2010}, the substructure velocity required to
produce such displacement
\citep{Milosavljevic2007,Springel2007,Mastropietro20008} and its
abundance in large N-body cosmological simulations
\citep{Hayashi2006,Lee2012, Thompson2012}. It has also been used to
constrain the dark matter particle self-interaction cross section and
to explore possible extensions to the concordance cosmological model
\citep{Farrar2007,Lee2010}.

Since then, other examples of Bullet-like systems have been found; 
MACS J0025.4-1222 \citep{Bradac2008}, Abell 2744 \citep{Merten2011},
DLSCL J0916.2+2951 \citep{Dawson2012}, ZwCl 1234.0+02916
\citep{Dahle2013}. Recently \citep{Gastaldello} observed a DM-gas
displacement of $87\pm 14$\hkpc and a DM-DM separation of $186\pm
30$\hkpc in \bullg, a low mass cluster system with a total
mass $2.4\pm 0.6 \times 10^{14}\Msun$ found in the the Strong Lensing
Legacy Survey (SL2S) sample \citep{Cabanac2007,More2012}.

Using a Sheth-Mo-Tormen mass function at $z=0$, one can estimate that
systems around this mass are $\sim 65$ times more abundant than massive
clusters in the mass range of the Bullet Cluster $>10^{15}$\hMsun
\citep{Sheth2001,hmfcalc}. This  should open up the possibility of
finding Bullet-like groups in large numbers to test
$\Lambda$CDM. However, a larger abundance of small mass systems has to
be weighted by the probability of being in a merger and  presenting a
large displacement between the DM and gaseous components.  These two
conditions (merger rates, maximum possible displacement) are a funcion
of DM halo mass in $\Lambda$CDM cosmologies.  A detailed statistical
study to estimate the DM-gas displacements of bullets has been
performed for clusters \citep{ForeroRomero2010} but not for lower mass
systems.

@@

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250\hMpc comoving on a side \citep{2011ApJ...740..120K}. The
simulation uses the ART code \citep{Kravtsov1997} to follow the
evolution of a dark matter density field from $z=80$ to $z=0$ sampled
with $2048^3$  particles. The cosmology used  corresponds to  the
spatially flat concordance model with the following parameters:  the
density parameter for matter (dark matter and baryons)
$\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the slope of
the primordial power spectrum $n=0.95$ and the amplitude of mass
density fluctuations (at redshift z$=$0) $\sigma_8=0.82$.  These
cosmological parameters are consistent with the nine-year Wilikinson
Microwave Anisotropy Probe (WMAP) results \citep{hinshaw_etal13}. A
detailed presentation of the simulation can be found in
\citet{2011ApJ...740..102K}.

@@

All the raw data used in this Letter are available through the
Multidark database \footnote{\url{www.multidark.org}}
\citep{Riebe2013}.  Furthermore, in order to facilitate the
reproducibility and reuse of our results we have made available all
the data and the source code available in a public
repository \footnote{\url{https://github.com/Fernandez-Trincado/Bullet\_Groups-2014}}.

@@

Each one of the pairs (host \& sub-halo) is considered as a
potential bullet system and is kept for the analysis described in the
next Section  by using the sub-halos as the tracer of the sub-dominant
dark matter clump in the merging cluster, i.e. the bullets.

@@

Bullet-like configurations have two dominant dark matter structures:
the host halo and the dominant sub-halo. We describe the kinematics of
this configuration by the  position and velocity vectors of the
sub-halo in a frame of reference where the main halo is at rest; thus
$\vec{v{}}=\vec{v}_{\rm sub}-\vec{v}_{\rm host}$ and
$\vec{r{}}=\vec{r}_{\rm sub}-\vec{r}_{\rm host}$, where the subscripts $host$
and $sub$ refer to the host and sub-halo, respectively. The positions
$\vec{r}$ correspond to the minium of potential and the velocities
$\vec{v}$ are computed as an average of the bulk velocities of all
simulation particles in the structure of interest.

@@

In this Letter we work with two quantities that could be inferred
from observations of Bullet-like systems. The projected distance
between two dominant DM clumps, $d_{\rm 2D}$, and the projected distance
between the DM and the gas clumps, $d_{\rm 2D}^{\rm bar}$. The
projection is computed along the same axis for all halos at all
redshifts, we present the result of the projection along the $z$-axis
of the simulation.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_1.pdf}
\end{center}
\caption{
  Integrated probability distribution for the displacement between the
  center of the host halo and its dominant sub-halo at all redshifts
  $z=0$, $0.25$, $0.5$ and $1.0$. The left panel
  shows the results in terms of the physical displacements while the
  right panel shows the displacements normalized by the virial radius
  of the host halo. The continuous (dashed) line corresponds to the halos in the
  cluster (group) sample.
  The vertical lines show the mean value and uncertainties ($186\pm
  30\hkpc$) in the  separation between the two dark matter clumps
  estimated in \citet{Gastaldello} for the \bullg. Between $40\%$
  to $60\%$ of the groups show a displacement equal or larger than
  this observational benchmark. This fraction rises to $60\%$ and
  $80\%$ in clusters.}
\label{fig:displacement}
\end{figure*}

Figure \ref{fig:displacement} presents the integrated
probability distribution for the DM-DM displacements, $d_{\rm 2D}$. The
left panel shows the displacement in physical units and the right
panel as a fraction of the virial radius of the host halo.  The panel
with the projected 2D physical displacements also shows a vertical
stripe with the estimated displacement for the bullet group reported
by \cite{Gastaldello}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_2.pdf}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$. The left
  panel corresponds to clusters and the right panel to groups. The data
  used to construct this histogram includes halos at all
  redshifts. This approximation to increase the statistical power of
  the sample is justified because the redshift evolution of this plot
  is weak.} 
\label{fig:geometry}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_3.pdf}
\end{center}
\caption{Integrated probability distribution for the estimated
  DM-gas displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and gas. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure*}

@@

In this Letter we estimated the fraction of galaxy groups and clusters
in a $\Lambda$CDM cosmology that could present observational features
associated to a Bullet-like event. This is motivated by the recent
observational results of \cite{Gastaldello} where a system (\bullg)
on the mass range $1\times 10^{14}$\hMsun\ and velocity dispersion
$650$\kms was reported to feature a displacement between its baryonic
(gas) and dark matter components.

@@

Our main result is that a fraction of $40\%$-$60\%$ of the halos
in the group sample presents DM-DM displacement equal or larger than the
observed displacement for \bullg. For halos in the cluster sample this
fraction increases to $60\%$-$80\%$. We also derived an
estimate for the DM-baryon displacement. In the group sample
$0.1\%$-$1.0\%$ of the halos show a displacement equal or larger than
the  measurements of \bullg\ by \cite{Gastaldello}; in the cluster
sample this fraction rises to $4\%$-$10\%$.

For the case of \bullg\ a fair comparison is achieved against our cluster
sample which has statistics dominated by objects of similar mass. In
a rough comparison using the observational criteria \citep{Foex2013,
  Gastaldello} we find that the relative number of Bullet-like systems
(large DM-gas displacement) with respect to a general sample of
multi-modal systems (large DM-DM displacements) is consistent with
observations; both are in the range $\sim 7\%$.

Using the same criteria we find that in the simulation there are
$\sim 6.0\times 10^{-7}$ Mpc$^{-3}$ groups similar to \bullg. This
number density is three times larger than the computed value for
clusters. This opens up a new observational possibility with surveys
such as SL2S that target a large number of groups and estimate its
multi-modal nature from lensing analysis \citep{Foex2013}. An approach
that can be further exploited with upcoming lensing surveys (e.g. with
 he {\it Euclid} satellite) and pushes for X-ray surveys with higher
sensibility: chances are larger to find bullets in systems with lower
X-ray luminosities.
